---
title: "HW3"
author: "Chenhui Yan"
date: "2024-10-02"
output: github_document
---
```{r}
library(tidyverse)
library(readxl)
library(devtools)
```
# Problem 1
## Part1: Data Cleaning:

* Retain variables: line, station, name, station latitude, longitude, routes served, entry, vending, entrance type, and ADA compliance.

* Convert entry to logical: The entry column has YES or NO values, so you'll need to convert it into a logical TRUE/FALSE format. You can use functions like ifelse() or case_when() to accomplish t

```{r}
# Read and clean the data
nyc_subway_data = read_csv("hw2_prob1_dataset.csv") %>% 
  janitor::clean_names() %>% 
  select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) %>% 
  mutate(entry = case_when(
    entry == "YES" ~ TRUE,
    entry == "NO"  ~ FALSE
  )) # Convert entry to logical TRUE/FALSE

# View the cleaned data
nyc_subway_data

```
## Part2: A brief summary of the dataset:

The cleaned dataset contains variables such as subway line, station name, latitude/longitude, routes served, entry status, vending, entrance type, and ADA compliance. To clean the data, I standardized column names using clean_names(), selected only relevant columns with select(), and converted the entry variable from character ("YES"/"NO") to logical (TRUE/FALSE) using case_when(). The resulting dataset has `r nrow(nyc_subway_data)` rows and `r ncol(nyc_subway_data)` columns, and it is tidy since each variable is a column, each observation is a row, and each value is a cell.