---
title: "HW3"
author: "Chenhui Yan"
date: "2024-10-02"
output: github_document
---
```{r}
library(tidyverse)
library(readxl)
library(devtools)
```
# Problem 1
## Part1: Data Cleaning:

* Retain variables: line, station, name, station latitude, longitude, routes served, entry, vending, entrance type, and ADA compliance.

* Convert entry to logical: The entry column has YES or NO values, so you'll need to convert it into a logical TRUE/FALSE format. You can use functions like ifelse() or case_when() to accomplish t

```{r}
# Read and clean the data
nyc_subway_data = read_csv("hw2_prob1_dataset.csv") %>% 
  janitor::clean_names() %>% 
  select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) %>% 
  mutate(entry = case_when(
    entry == "YES" ~ TRUE,
    entry == "NO"  ~ FALSE
  )) # Convert entry to logical TRUE/FALSE

# View the cleaned data
nyc_subway_data

```
## Part2: A brief summary of the dataset:

The cleaned dataset contains variables such as subway line, station name, latitude/longitude, routes served, entry status, vending, entrance type, and ADA compliance. To clean the data, I standardized column names using clean_names(), selected only relevant columns with select(), and converted the entry variable from character ("YES"/"NO") to logical (TRUE/FALSE) using case_when(). The resulting dataset has `r nrow(nyc_subway_data)` rows and `r ncol(nyc_subway_data)` columns, and it is tidy since each variable is a column, each observation is a row, and each value is a cell.

## Part3: Answer the following questions using these data:
### How many distinct stations are there? 
```{r}
distinct_station = 
   nyc_subway_data %>%
  distinct(line, station_name) %>%
  nrow()
```
There are `r distinct_station` distinct stations in this dataset. 

### How many stations are ADA compliant?
```{r}
ada_compliant =
  nyc_subway_data %>% 
  filter(ada == TRUE) %>% 
  distinct(line, station_name) %>% 
  nrow()
```
There are `r ada_compliant` ADA compliant stations. 

### What proportion of station entrances / exits without vending allow entrance?
```{r}
proportion_no_vending_allows_entry = 
  nyc_subway_data %>%
  filter(vending == "NO") %>%  # Filter rows where there is no vending
  summarise(proportion = mean(entry == TRUE)) %>%  # Calculate the proportion where entry is TRUE
  pull(proportion)

# Output the result
proportion_no_vending_allows_entry
```
37.7% of station entrances / exits without vending allow entrance

## Part4: Reformat data so that route number and route name are distinct variables.
```{r reformat routes}
nyc_subway_data_refored = 
  nyc_subway_data %>%
  mutate(across(starts_with("route"), as.character)) %>%  # Ensure all route columns are characters
  pivot_longer(
    cols = starts_with("route"), # Specify the columns for the routes
    names_to = "route_number",
    values_to = "route") %>%
  drop_na(route)# Drop rows where route is NA
nyc_subway_data_refored

```

### How many distinct stations serve the A train?*
```{r distinct stations A train}
nyc_subway_data_refored %>% 
  filter(route == "A") %>% 
  distinct(line, station_name) %>% 
  nrow()
```
There are distinct stations serve the A train. 


### How many are ADA compliant of the stations that serve the A train?*
```{r ADA compliant A train}
nyc_subway_data_refored %>% 
  filter(ada == TRUE) %>%  
  filter(route == "A") %>% 
  distinct(line, station_name) %>% 
  nrow()
```
Of the 60 stations that serve the A train,17 have at least one ADA compliant entry or exit.

# Problem 2
## Part1. Data Cleaning for Mr trash wheel sheet
```{r}
# Specify sheet and omit non-data entries (skip rows and specify range)
mr_trash_wheel = read_excel("hw2_prob2_datas.xlsx",
                            sheet = "Mr. Trash Wheel", 
                            skip = 1,  # Skip the first row if it's non-data or a note
                            range = "A2:N653")  # Adjust the range based on the actual data layout
# Omit rows that do not include dumpster-specific data
mr_trash_wheel_cleaned = mr_trash_wheel %>%
  filter(!is.na(Dumpster)) %>%  # Keep only rows where the 'dumpster' column is not NA
  janitor::clean_names() %>%
  select(dumpster:homes_powered) %>% 
  mutate(sports_balls = round(sports_balls, 0)) %>% 
  mutate(sports_balls = as.integer(sports_balls)) %>% 
  mutate(year, year = as.numeric(year))
mr_trash_wheel_cleaned

```
